Toward A Non-Atomic Algorithmic Lov\'{a}sz~ Local Lemma

### 2015-05-28 paper/nonatomic/paper.tex
Some work on an expanded version of the 270 survey.

@@

@@

\abstract{The \lovasz Local Lemma (LLL) \cite{erdos1975problems} is a probabilistic method for proving the existence of elements of a $\sigma$-algebra $(\Omega, \Sigma)$ that miss all of a set of events $\mathcal{A}$.  The method works when a probability measure can be assigned to $\Sigma$ so that the marginal probability of the events is low and the their statistical dependency graph is sparse.  In the less abstract \emph{variable version} of the LLL problem, the space is a product space of, say, $n$ independent variables and each event corresponds to a binary function of some of the variables.  In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} made the LLL method constructive for variable LLL problems, giving a surprisingly simple expected-polynomial-time algorithm to find variable assignments that make no events happen.  They used a novel entropy compression argument, proving that their algorithm could be used to compress random bits if it took more than a certain amount of expected time to find a solution.  Since then, there has been a flurry of follow-up work.  In this paper, we briefly review the Moser-Tardos result and a recent generalization by Achlioptas and Iliopoulos \cite{achlioptas2014random}).  Then we propose a generalization of the entropy compression argument to new ``non-atomic'' settings using \emph{relative entropy compression}.}

\section{Introduction}
The \lovasz Local Lemma (LLL) is a probabilistic method for proving the existence of elements of a probability space $\mathcal{P} = (\Omega, \Sigma, \mu)$ that miss all of a set of events $\mathcal{A} = \{A_1, \cdots, A_m\}$, the search for which we call an ``LLL problem''.  It was introduced by \lovasz and Erd\"{o}s in 1975 \cite{erdos1975problems}; the proof is a fairly straightforward application of the probabilistic method \cite{alon2004probabilistic}.  The LLL has many applications in combinatorial search problems.  A useful restricted version of the problem is the ``variable version,'' in which $\mathcal{P}$ is a product space of $n$ independent random variables $X = (X_1, \cdots, X_n)$.  (We will sometimes write $A_i(X)$ for the indicator that event $i$ happens when the variables take value $X$.)  For example, we can use the LLL to prove the existence of solutions to certain \ksat problems by imagining a probability space on $n$ random variables in which each event $A_j$ occurs if a corresponding disjunctive clause $C_j \in \mathcal{C} = \{C_1, \cdots, C_m\}$ of $k$ variables is false.  In that setup, ``missing all events'' corresponds to satisfying all clauses.  Not every \ksat problem has a satisfying solution, so we should expect the conditions of the LLL to be restrictive.  We will come back to the \ksat example several times, since it is a very straightforward application.

A seminal paper by Beck \cite{beck1991algorithmic} showed that the LLL could be made \emph{algorithmic} -- that is, when the LLL conditions hold, certain algorithms will \emph{find} a solution quickly.  This started a literature around the ``constructive'' or ``algorithmic'' LLL.  Beck's result required stronger conditions than the non-constructive LLL and worked only for particular problems (like \ksat).  In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} published a breakthrough meta-algorithm for finding solutions to any variable LLL problem under the same conditions as the non-constructive LLL.  In its simplest form, the MT algorithm is strikingly simple:

\begin{algorithm}[H]
\caption{The Moser-Tardos (MT) algorithm, in simple iterative form.}
\label{alg:mt}
\begin{algorithmic}
\State Initialize the $X_i$ IID uniform.
\While{Any of the $A_j$ happen under the current value of $X$}
  \State Let $A_j$ be an arbitrary event that currently happens.
  \State Resample each $X_i \in \operatorname{vbl}(A_j)$ IID uniform.
\EndWhile
\end{algorithmic}
\end{algorithm}

For example, specialized to \ksat, the MT algorithm becomes:

\begin{algorithm}[H]
\caption{The Moser-Tardos algorithm, specialized to \ksat.}
\label{alg:mt-ksat}
\begin{algorithmic}[1]
\State Initialize the $X_i$ IID uniform Bernoulli.
\While{Any clause is violated by $X$}
  \State Let $C_j$ be an arbitrary clause that is violated.
  \State Resample each $X_i \in \operatorname{vbl}(C_j)$ IID uniform Bernoulli.
\EndWhile
\end{algorithmic}
\end{algorithm}

Moser and Tardos prove that their algorithm finishes in expected polynomial time under the \emph{standard LLL condition} $e p (d+1) < 1$, where
\begin{description}
  \item[$e$] is the base of the natural logarithm;
  \item[$p$] is the maximum marginal probability of any event when $X \hasDist \operatorname{Unif}(\Omega)$; and
  \item[$d$] is the maximum degree of any event in the \emph{dependency graph} on events, in which two events are connected by an edge if they are statistically dependent when $X \hasDist \operatorname{Unif}(\Omega)$ .
\end{description}

After the Moser and Tardos paper, interest in the subject has only increased; follow-up work has given new algorithms, new problem domains, new applications, and a range of sufficient conditions weaker than the LLL condition.

\subsection{This paper}
In this paper, we first review the result by Moser and Tardos, using a more illuminating proof by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that directly highlights the role of entropy compression.  In section \ref{sec:perfect}, we review the extension of the algorithmic LLL to non-variable problems by Achlioptas and Iliopoulos \cite{achlioptas2014random}.  Using these two ideas, in section \ref{sec:nonatomic} we sketch an argument that suggests the entropy compression argument could be extended to novel LLL-like applications in which the state vector of an algorithm cannot be uniquely recovered.  Achlioptas and Iliopoulos call such algorithms (or rather, technically, their associated random walk graphs) ``non-atomic''.  The original work in this paper is preliminary and incomplete, and we look forward to improving it in the near future.

@@

\begin{algorithm}[H]
\caption{A more complicated recursive version of the Moser-Tardos (MT) algorithm.  Here $\Gamma(i)$ denotes the neighbors of event $i$ in the event dependency graph.}
\label{alg:mt-recursive}
\begin{algorithmic}
\State Initialize the $X_i$ IID uniform.
\ForAll{$i \in [m]$}
  \Call{Fix}{i}
\EndFor

\Function{Fix}{i}
  \If{$! A_i(X)$}
    \Return
  \EndIf
  \State Resample each $X_i \in \operatorname{vbl}(A_i)$ IID uniform.
  \ForAll{$j \in [m]: j \in \Gamma(i), A_j(X)$}
    \Call{Fix}{j}
  \EndFor
\EndFunction

@@

We would like to prove that this algorithm terminates in polynomial time when the LLL condition is met.  We will briefly review an argument by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that emphasizes the entropy compression aspect of the proof.  The argument is specialized to \ksat to make the exposition simpler.

The general idea is to imagine running algorithm \ref{alg:mt-recursive} until \texttt{Fix} is called $t$ times.  If $i(s)$ denotes the argument to \texttt{Fix} on iteration $s$, then after $t$ calls, $\sum_{j = 1}^{t} \operatorname{vbl}(A_{t(j)}) = k t$ binary variables have been resampled independently using some random source, and the initial assignment used $n$ bits.  Together, the algorithm state vector, which we call $Z$, encodes $n + k t$ purely random bits.  We will use the assumption $2^C_0 p d < 1$ to give a method to encode the algorithm state in an alternative way $Y$, which we call the ``alternative information''.  (We will use $C_0 = 3$, though the proof is possible under the weaker condition $e p (d + 1) < 1$.)  When $t$ is large, $Y$ will contain fewer bits than $Z$, which contradicts Shannon's source coding theorem \cite{shannon2001mathematical} (it is impossible to encode purely random bits with a scheme that uses fewer bits in expectation).

We note first that, given an innocuous assumption about the problem (which is true for \ksat and can be made true for other problems), there is a 1-to-1 correspondence between the pair $((A_{t(i)})_{i=1}^{t}, \sigma_{t+1})$ and $Z$.  Here $\sigma_t$ denotes the state of $X$ before call $t$.  The assumption we need is that each event $A_i$ happens for only one setting of $\operatorname{vbl}(A_i)$.  For example, in \ksat, a disjunctive clause can only be violated by one setting of its variables.  In section \ref{subsec:trivial} we will explain why this assumption is innocuous in general.  Under the assumption, there is an easy inductive argument: If $\sigma_{i+1}$ and $A_{t(i)}$ are known, then $\sigma_i$ must equal $\sigma_{i+1}$ outside $\operatorname{vbl}(A_{t(i)})$, and on $\operatorname{vbl}(A_{t(i)})$ it must equal the one violating setting, since it was chosen to be \texttt{Fix}ed.  But $\sigma_{t+1}$ and all the events are known, so the whole state vector $Z$ (including the initial state) is determined.

It remains to encode $(A_{t(i)})_{i=1}^{t}$ efficiently given $\sigma_{t+1}$.  Note that we cannot simply encode the events using $\log m$ bits, since this would not give us the bound we want.  Instead we will use a structure called (by Moser and Tardos) a ``witness forest.''  In this forest, each event called directly by the outer loop of the algorithm (and which actually causes a resample) is a root, and an event called inside $\texttt{Fix}(i)$ is a child of event $A_i$.  For \ksat, $p = 2^{-k}$, so by the assumption $2^C_0 p d < 1$, each event is independent of all but at most $2^{k-C_0}$ other events.  So if an event is a child, it can be encoded using at most $2^{k-C_0}$ bits.  Each event is either a root or not, so the roots can be encoded using $m$ bits.  Finally, to encode each tree, we can use an extra bit per element to denote the end of the tree.

So we can encode $Y$ using at most $n + 2m + (k - C_0 + 1)t$ bits, and we can determine $X$ from $Y$, so by Shannon's source coding theorem
\seqn
  n + m + (k - C_0 + 1)t \leq n + k t .
\eeqn

This is violated if $t \geq \frac{m}{C_0 - 1}$, so the algorithm must finish (in expectation) before then.  (In particular, it must finish \emph{eventually}, so there must exist a satisfying solution, and this argument is another proof of the LLL.)

\section{The algorithmic LLL on general sample spaces}
\label{sec:perfect}
Thus far, we have seen only a variable-version LLL problem in which our probability space was a product of $n$ variables.  The MT algorithm requires explicit access to those variables, and it is restricted to handling every event in the same way -- by resampling all of its variables.  Both assumptions are fairly restrictive.  The sample space for some combinatorial problems is better represented in other forms, like permutations \cite{achlioptas2014random}.  And contra the second assumption, many local search algorithms for \ksat (e.g. \texttt{WalkSAT} \cite{papadimitriou1991selecting}) are similar to MT, but will flip only a smaller number of variables on each iteration.  Achlioptas and Iliopoulos \cite{achlioptas2014random} provide an elegant framework that allows the probability space to be arbitrary and the algorithm to choose actions more flexibly, for which variable-version problems and algorithms are a special case.  Examining this generalization is instructive.

Again take our $\sigma$-algebra to be $(\Omega, \Sigma)$, but this time without a product structure and without a measure $\mu$.  Our set of events, which we will now call ``flaws'', are the same as before, but now there is no notion of $\operatorname{vbl}(A_i)$.  Let $U(\sigma)$ denote the set of flaws that happen under state $\sigma$.  Again, our goal is to find any element $\sigma^* \in \Omega$ such that none of the events happen; we call such elements with $U(\sigma^*) = \emptyset$ ``flawless objects.''  Consider any randomized iterative algorithm whose random state is Markovian and stationary, which attempts to find a flawless object.  (The MT algorithm is an example of such an algorithm.)  We can always encode such an algorithm in a random walk graph $\mathcal{T}$ on $\Omega$.  Now let us try to write down the properties of this graph that are really needed for something like the proof in section \ref{sec:lll} to apply.

The first requirement is ``accountability of action.''  In Tao's proof, we needed the fact that the flaw (event) $A_t$ addressed on iteration $t$ uniquely determined the values of $\operatorname{vbl}(A_t)$ in the previous state $\sigma_t$.  We proved this by assuming, by induction, that we knew the values of the variables in $\sigma_{t+1}$.  Given the identity of the flaw, we knew that $\sigma_t$ and $\sigma_{t+1}$ differed only on the variables in $\operatorname{vbl}(A_t)$ differed.  Further, since each flaw could be violated by only one configuration of its variables, we knew what that configuration must have been in $\sigma_t$.  Without variables, we need a different tack to get a correspondence between flaw vectors and algorithm state vectors.  We will still assume that the algorithm ``addresses'' a single flaw on each iteration; given a current state $\sigma_t$ and a flaw to address, the algorithm draws from some distribution to find its next action.  To achieve accountability, we assume that, for every (flaw, state) pair $(A_{t}, \sigma_{t+1})$, there is a unique previous state $\sigma_t$.  In terms of the random walk graph, this means that if we label each weighted edge in the graph with the flaw we addressed at that step, each node has at most one incoming edge for each flaw.  This requirement is called ``atomicity.''

>>>>

@article{formentin2009symmetric,
  title={A symmetric entropy bound on the non-reconstruction regime of Markov chains on Galton-Watson trees},
  author={Formentin, Marco and K{\"u}lske, Christof},
  journal={arXiv preprint arXiv:0903.2962},
  year={2009}

@@

@article{prufer1918neuer,
  title={Neuer beweis eines satzes {\"u}ber permutationen},
  author={Pr{\"u}fer, Heinz},
  journal={Arch. Math. Phys},
  volume={27},
  pages={742--744},
  year={1918}
@inproceedings{neville1953codifying,
  title={The codifying of tree-structure},
  author={Neville, EH},
  booktitle={Mathematical Proceedings of the Cambridge Philosophical Society},
  volume={49},
  number={03},
  pages={381--385},
  year={1953},
  organization={Cambridge Univ Press}
@article{otter1948number,
  title={The number of trees},
  author={Otter, Richard},
  journal={Annals of Mathematics},
  pages={583--599},
  year={1948},
  publisher={JSTOR}
@inproceedings{jacobson1989space,
  title={Space-efficient static trees and graphs},
  author={Jacobson, Guy},
  booktitle={Foundations of Computer Science, 1989., 30th Annual Symposium on},
  pages={549--554},
  year={1989},
@book{drmota2009random,
  title={Random trees: an interplay between combinatorics and probability},
  author={Drmota, Michael},
  publisher={Springer Science \& Business Media}

@@

\abstract{The \lovasz Local Lemma (LLL) \cite{erdos1975problems} is a probabilistic method for proving the existence of elements of a $\sigma$-algebra $(\Omega, \Sigma)$ that miss all of a set of events $\mathcal{A}$.  The method works when a probability measure can be assigned to $\Sigma$ so that the marginal probability of the events is low and the their statistical dependency graph is sparse.  In the less abstract \emph{variable version} of the LLL problem, the space is a product space of, say, $n$ independent variables and each event corresponds to a binary function of some of the variables.  In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} made the LLL method efficiently constructive for variable LLL problems, giving a surprisingly simple expected-polynomial-time algorithm to find variable assignments that make no events happen.  They used a novel entropy compression argument, proving that their algorithm could be used to compress random bits if it took more than a certain amount of expected time to find a solution.  Since then, there has been a flurry of follow-up work.  In this paper, we briefly review the Moser-Tardos result and a recent generalization by Achlioptas and Iliopoulos \cite{achlioptas2014random}).  Then we propose a generalization of the entropy compression argument to new ``non-atomic'' settings using \emph{relative entropy compression}.}

\section{Introduction}
The \lovasz Local Lemma (LLL) is a probabilistic method for proving the existence of elements of a probability space $\mathcal{P} = (\Omega, \Sigma, \mu)$ that are flawless -- elements that miss all of a set of bad events (``flaws'') $\mathcal{A} = \{A_1, \cdots, A_m\}$.  We call any search for such elements an ``LLL problem''.  It was introduced by \lovasz and Erd\"{o}s in 1975 \cite{erdos1975problems}; the proof is a fairly straightforward application of the probabilistic method \cite{alon2004probabilistic}.  The LLL has many applications in combinatorial search problems.

A less abstract (though somewhat restricted) version of the problem is the ``variable version,'' in which each element of $\Omega$ can be expressed as a vector of $n$ variables, and those variables are independent.  More formally, in a variable-LLL problem, $\mathcal{P}$ is a product of $n$ probability spaces on which we define $n$ corresponding independent random variables $(X_1, \cdots, X_n) = X$ product measure $\prod_{j = 1}^{n} \mu_j  = \mu$.  Each $A_i$ is a set of values of $X$ that exhibit some abstract ``flaw.''  In a slight abuse of notation, we will write an event $A_i$ as a function of $X$ that returns $1$ when the corresponding event happens and $0$ otherwise.  As a function, $A_i$ may be functionally dependent on only a subvector of $X$, and we will write $\operatorname{vbl}(A_i) \subseteq [n]$ for the indices on which function $A_i$ depends.  Finally, following \cite{achlioptas2014random}, we will denote by $U(X)$ the events that happen under $X$, or more colloquially the ``flaws of $X$.''

For example, we can use the LLL to prove the existence of solutions to certain \ksat problems.  In applying the LLL to \ksat, we imagine a probability space on $n$ independent uniform random binary variables in which each event $A_j$ occurs if a corresponding disjunctive clause $C_j \in \mathcal{C} = \{C_1, \cdots, C_m\}$ of $k$ of the variables is false under the variable setting $X$.  (We denote by $\operatorname{vbl}(C_j)$ the set of $k$ variables of clause $C_j$.)  A flawless element $X$ is just one that satisfies all clauses.  Not every \ksat problem has a satisfying solution, so we should expect the conditions of the LLL to be restrictive.  We will come back to the \ksat example several times, since it is a very straightforward application.

The original LLL does not tell us how to \emph{construct} a solution when its conditions are met.  There is an obvious Las Vegas algorithm for finding a solution -- since a solution exists, just sample from $\mathcal{P}$ until one is found.  In fact, the LLL guarantees more than the existence of a solution; it provides a lower bound on $\mu(\cap_{i \in [m]} \bar{A_i})$.  Unfortunately, this lower bound can be exponentially small, so there is no guarantee that the Las Vegas algorithm is efficient.  But a seminal paper by Beck \cite{beck1991algorithmic} showed that the LLL \emph{could} be made algorithmic.  That is, when the LLL conditions hold, certain algorithms will find a solution quickly.  This started a literature around the ``constructive'' or ``algorithmic'' LLL.  Beck's result required stronger conditions than the non-constructive LLL and worked only for particular problems (like \ksat).  In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} published a breakthrough meta-algorithm for finding solutions to any variable LLL problem under the same conditions as the non-constructive LLL.  In its simplest form, the MT algorithm is strikingly simple:

\begin{algorithm}[H]
\caption{The Moser-Tardos (MT) algorithm, in simple iterative form.}
\label{alg:mt}
\begin{algorithmic}
\Function{Solve}{$\mathcal{P}$, $\mathcal{A}$}
  \State Initialize $X \hasDist \mu$.
  \While{$X$ has a flaw}
    \State Let $A_j$ be an arbitrary flaw of $X$.
    \State Resample each $X_i \in \operatorname{vbl}(A_j) \hasDist \mu_i$.
  \EndWhile
\EndFunction
\end{algorithmic}
\end{algorithm}

For example, specialized to our formulation of \ksat, the MT algorithm becomes:

\begin{algorithm}[H]
\caption{The Moser-Tardos algorithm, specialized to \ksat.}
\label{alg:mt-ksat}
\begin{algorithmic}[1]
\Function{Solve}{$n$, $\mathcal{C}$}
  \State Initialize $X_1, \cdots, X_n$ i.i.d. uniform Bernoulli.
  \While{Any clause in $\mathcal{C}$ is violated by $X$}
    \State Let $C_j$ be an arbitrary clause that is violated.
    \State Resample each $X_i \in \operatorname{vbl}(C_j)$ i.i.d. uniform Bernoulli.
  \EndWhile
\EndFunction

Moser and Tardos prove that their algorithm finishes in expected polynomial time under the \emph{standard LLL condition} $e p (d+1) < 1$, where
\begin{description}
  \item[$e$] is the base of the natural logarithm;
  \item[$p$] is the maximum marginal probability of any flaw under distribution $\mu$, $\sup_{i \in [m]} \mu(A_i)$; and
  \item[$d$] is the maximum degree of any flaw in the \emph{dependency graph} on flaws, in which two flaws are connected by an edge if they are statistically dependent when $X \hasDist \mu$.  (We will denote by $\Gamma(i)$ the neighborhood of $i$, the set of indices of flaws that neighbor flaw $i$ in this dependency graph.  $\Gamma^{+}(i) = \Gamma(i) \cup \{i\}$ is the inclusive neighborhood.)  Then $d = \max_{i \in [m]} |\Gamma(i)|$ and $d+1 = \max_{i \in [m]} |\Gamma^{+}(i)|$.
\end{description}

Their proof uses a novel \emph{entropy compression} argument.  They show that the MT algorithm must terminate in a certain number of steps (on average) or it would be possible to represent the algorithm's state trajectory using fewer random bits than were used to generate it, which is impossible.  Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} make this argument more explicit.  Notice that taking $\mu$ to be the uniform distribution maximizes the entropy created at each step of the algorithm (at least on finite spaces), so it is easiest to make an entropy compression argument in that case.

After the Moser and Tardos paper, interest in the subject has only increased; follow-up work has given new algorithms, new problem domains, new applications, and a range of sufficient conditions weaker than the LLL condition.  Our interest, and the subject of this paper, is in strengthening the Moser-Tardos result from the \emph{variable} version of the problem to more general versions, under analogous conditions.  The key proof idea is taking the entropy compression argument seriously. %FIXME: Language.  Sounds arrogant.

\subsection{This paper}
In this paper, we first review the result by Moser and Tardos, using the more illuminating proof by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that directly highlights the role of entropy compression.  In section \ref{sec:perfect}, we review an extension of the algorithmic LLL to non-variable problems by Achlioptas and Iliopoulos \cite{achlioptas2014random}.  Achlioptas and Iliopoulos give a general condition on \emph{algorithms} rather than problems, and bring us partway to a general constructive LLL, but they require that the space $\Omega$ is finite, and this is the restriction we will remove.  In section \ref{sec:examples} we give a series of examples suggesting the entropy compression argument could be fruitfully extended to novel LLL-like applications in which the state vector of an algorithm cannot be uniquely recovered.  Achlioptas and Iliopoulos call such algorithms (or, technically, their associated random walk graphs) ``non-atomic''.  By removing the atomicity assumption, we can develop a more general constructive LLL.  In section \ref{sec:nonatomic}, we give a formal proof.  %FIXME: Language.

@@

\begin{algorithm}[H]
\caption{A more complicated recursive version of the Moser-Tardos (MT) algorithm.  This version replaces the \emph{arbitrary} order of resampling in algorithm \ref{alg:mt} with a fixed order that resamples neighborhoods of events in the dependency graph.}
\label{alg:mt-recursive}
\begin{algorithmic}
\Function{Solve}{$\mathcal{P}$, $\mathcal{A}$}
  \State Initialize $X \hasDist \mu$.
  \ForAll{$i \in [m]$}
    \Call{Fix}{i}
  \EndFor
  \Return $X$
\EndFunction

\Function{Fix}{i}
  \If{$! A_i(X)$}
    \Return
  \EndIf
  \State \Call{Resample}{i}
  \ForAll{$k \in [m] \texttt{s.t.} k \in \Gamma^{+}(i) \texttt{ and } A_k(X) = 1$}
    \Call{Fix}{j}
  \EndFor
\EndFunction

\Function{Resample}{i}
  \State Resample each $X_j \in \operatorname{vbl}(A_i) \hasDist \mu_j$.
\EndFunction

@@

An an introduction to later parts of this paper, we will give a proof that this algorithm terminates in polynomial time when the LLL condition is met.  Our argument mostly follows one by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that emphasizes the entropy compression aspect of the proof.  The argument is specialized to \ksat to make the exposition simpler, but we will later discuss the generalization to arbitary LLL problems.  We will prove that the specialization of algorithm \ref{alg:mt-recursive} to \ksat terminates before $m/2$ calls to \texttt{Resample}. %FIXME: Might be extra O(1) terms.
Recall that in the specialization to \ksat, $\mu$ is the uniform distribution on $\{0,1\}^n$-vectors and $A_i(X) = 1$ if clause $C_i$ is false.

%FIXME: This could do with division into subsections, since it is so long.

\subsection{A nice property of algorithm \ref{alg:mt-recursive}}
First, algorithm \ref{alg:mt-recursive} has a nice property that the simpler algorithm \ref{alg:mt} does not have, which motivates our switch to the more complicated version.  If \texttt{Fix}$(i)$ returns, $X$ cannot have flaw $A_i$, nor can it have any flaw it did not have before \texttt{Fix}$(i)$ was called.  The proof is by induction on the tree of calls to \texttt{Resample} caused by calling \texttt{Fix}$(i)$, which we (following Moser and Tardos) call a ``witness tree.''  In the witness tree (which we will be important later), each node is labeled (possibly nonuniquely) with a flaw index.  A call to \texttt{Fix}$(i)$ may cause a call to \texttt{Resample}$(i)$ and zero or more additional calls to \texttt{Fix} in the following loop.  There is a node in the tree for the call to \texttt{Resample}$(i)$, and that node has a child for each call to \texttt{Resample} caused by the calls to \texttt{Fix} in the loop.
%FIXME: Picture

The inductive argument is fairly simple.  If the tree is empty, then $X$ must not have flaw $A_i$, and no resampling occurred, so no additional flaws are added.  Otherwise, the variables $\operatorname{vbl}(A_i)$ were resampled and then calls to \texttt{Fix} were made for elements of clause $A_i$'s inclusive neighborhood (including $A_i$ itself) that were unsatisfied.  No other clauses could have become unsatisfied by this resampling, since clauses outside $\Gamma^{+}(i)$ do not have variables $\operatorname{vbl}(A_i)$.  By induction, these calls to \texttt{Fix} add no new flaws and satisfy each clause in $\Gamma^{+}(i)$.  So our call to \texttt{Fix}$(i)$ added no new flaws and ensured that $A_i$ was satisfied.
%FIXME: A little loose with the A_i and C_i notation here.
%FIXME: Should merge the witness tree description here, or break it into its own section.

This property implies that \ref{alg:mt-recursive} finds a satisfying solution if \texttt{Solve} returns at all, since it loops over all flaws and fixes each without introducing any new ones.  So it remains to show when (and how quickly) that happens.

We should emphasize that this fact is only \emph{convenient}, not necessary, for the following proof.  Algorithm \ref{alg:mt} enjoys many of the same guarantees as algorithm \ref{alg:mt-recursive}.

\subsection{The entropy compression argument}
The idea is to imagine running algorithm \ref{alg:mt-recursive} until \texttt{Resample} is called $t$ times.  We will give a scheme for compressing random bits -- which is impossible -- if $t$ is too large in expectation.  Let $\sigma_s$ denote the value of $X$ at the beginning of the $s$th call to \texttt{Resample}, and let $i(s)$ denote the argument to \texttt{Resample} on that iteration.  We will say that $A_{i(s)}$ was ``addressed'' on iteration $s$.  After $t$ calls, $\sum_{j = 1}^{t} \operatorname{vbl}(A_{t(j)}) = k t$ binary variables have been resampled independently using some random source.  The initial assignment used $n$ bits.  Let $Z \in \Omega^{t+1} = (\sigma_0, \sigma_1, \cdots, \sigma_t)$ denote the algorithm state vector.  Then $Z$ encodes $n + k t$ purely random bits.  We will use the assumption $2^C_0 p (d+1) < 1$ to give a method to encode the algorithm state in an alternative way $Y$, which we call the ``alternative information''.  (We will use $C_0 = 2$ and later $C_0 = \log_2 \frac{1}{0.3383}$, though the tightest constant is $C_0 = \log_2 e$.)  When $t$ is large on average, $Y$ will contain fewer bits than $Z$ on average.  That contradicts Shannon's source coding theorem \cite{shannon2001mathematical}, which formalizes the intuitive fact that it is impossible to encode purely random bits with a scheme that uses fewer bits in expectation.

We note first that, given an innocuous assumption about the problem (which is true for \ksat and can be made true for other problems), there is a 1-to-1 correspondence between the pair $((A_{t(i)})_{i=1}^{t}, \sigma_{t})$ and $Z$.  In other words, we assume that if we know the final return value and the sequence of flaws that were addressed, we know the whole sequence of states the algorithm passed through.  The assumption we need is that each flaw $A_i$ happens for only one setting of $\operatorname{vbl}(A_i)$.  For example, in \ksat, a disjunctive clause can only be violated by one setting of its variables.  In section \ref{subsec:trivial} we will explain why this assumption is innocuous in general.  Under the assumption, there is an easy inductive argument: If $\sigma_{i+1}$ and $A_{t(i)}$ are known, then $\sigma_i$ must equal $\sigma_{i+1}$ outside $\operatorname{vbl}(A_{t(i)})$, and on $\operatorname{vbl}(A_{t(i)})$ it must equal the one violating setting, since it was chosen to be \texttt{Resample}d.  But $\sigma_{t+1}$ and all the flaws are known, so the whole state vector $Z$ (including the initial state) is determined.
%FIXME: Include a picture.

It remains to encode $(A_{t(i)})_{i=1}^{t}$ efficiently.  Note that we cannot simply encode the addressed flaws using the natural encoding that takes $\log m$ bits, since this would lead us a different, much stronger condition than the LLL condition.  Instead we will use a structure called (by Moser and Tardos) a ``witness forest'' to hold instances in which flaws are addressed.  In this forest, there are two kinds of nodes:
\begin{description}
  \item[Roots:] Whenever the outer loop of the algorithm calls \texttt{Resample}$(i)$ and flaw $A_i$ currently happens, that flaw's variables are resampled, and then its neighbors in the dependency graph are potentially addressed.  We designate such outer-loop flaw-addressing instances as roots in the witness forest.  (When \texttt{Fix}$(i)$ is called but flaw $A_i$ does not currently happen, we add nothing to the forest.)
  \item[Children:] Whenever the inner loop in \texttt{Fix}$(i)$ calls \texttt{Fix}$(j)$, there must already be a node in the tree (either a root or, inductively, a child) corresponding to that flaw-addressing instance for flaw $i$.  When such a call results in a call to \texttt{Resample}$(j)$, we add a child to that node corresponding to flaw $A_j$.
\end{description}

For \ksat, $p = 2^{-k}$, so by the assumption $2^C_0 p (d+1) < 1$, each flaw is independent of all but at most $2^{k-C_0}$ other flaws, including itself.  So if an flaw-addressing instance is a child in the witness forest, it can be encoded using only $k-C_0$ bits.  Unfortunately, the argument is more complicated than that, because we must also encode the structure of the forest itself.  Encoding the identities and order of the roots is easy enough, since each flaw can appear at most once, and the order is by flaw index.  But encoding the individual trees is harder.  Given a complete graph on $t$ nodes, there are $t^{t-2}$ distinct spanning trees, so it takes $(t-2) \log_2 t$ bits to encode such trees \cite{prufer1918neuer,neville1953codifying}.  Since we also need to know the order of calls in the inner loop, and we can have potentially $m$ disjoint trees, the encoding length of a witness forest appears to be even greater than this.  If this were true, the average encoding length of a node would eventually become larger than $k$, ruining the proof.

However, we actually only need to encode the \emph{unlabelled structure} of each tree in the witness forest, and this is easier.  There is a simple encoding scheme, but let us add some intuition.  First, note that recording the order of calls in the inner loop is unnecessary, since the calls are made in a fixed order by flaw identity.  So if we are told only the identity of the root of each tree in the forest and the set of children of each node, we can reconstruct the order of each child set.  %FIXME: This is a bad explanation.  Needs pictures.
Second, we do not actually need to label the nodes of the tree.  We can just store the $k-C_0$-bit child identifiers in a some fixed order, for example the order in which a DFS visits nodes of the tree when children are ordered by their indices. %FIXME: Also needs a picture.
So the number of distinct witness trees to encode (ignoring the issue of encoding the roots, which we have already) is at most the number of \emph{unlabelled rooted trees} on $t$ nodes, which is much smaller than the number of labelled trees.  While there is no known exact formula for counting such trees, Otter \cite{otter1948number} gives an asymptotic estimate of
\seqn
  |\{T: T\texttt{ is an unlabelled tree}, |T| = n\}| \tilde O(1) (\frac{1}{0.3382\cdots})^t + O(1) .
\eeqn
Therefore (by Shannon's noisy channel coding theorem) there exists an encoding scheme for witness trees that uses only, say, $(\log_2 \frac{1}{0.3383}) t + O(1)$ bits.  Note that $\log_2 \frac{1}{0.3383}$ is only an upper bound on the number of required bits per iteration of the algorithm, since it assumes the maximum-entropy (uniform) distribution on unlabelled trees of size $n$, while the algorithm may generate witness trees with less entropy than that.

If we are willing to weaken the upper bound a bit, there is a very simple concrete encoding scheme that follows the algorithm execution.  First, we record the $k-C_0$-bit flaw data in the order in which \texttt{Resample} was called.  Then we record the history of the depth-first search executed in each tree by logging a $0$ whenever \texttt{Fix} is called inside \texttt{Fix}, and a $1$ whenever a call to \texttt{Fix} returns.  This uses at most $2$ bits per edge in the witness trees, thus at most $2t$ total bits.

So we can encode $Y$ using at most $n + m + (k - C_0 + 2)t$ bits, and we can determine $Z$ from $Y$, so by Shannon's source coding theorem
\seqn
  n + m + (k - C_0 + 1)t \leq n + k t .
\eeqn

This is violated if $t \geq \frac{m}{C_0 - 2}$, so the algorithm must finish (in expectation) before that many iterations.  We have also shown that the algorithm must finish in expectation before $\frac{m}{C_0 - \log_2 \frac{1}{0.3383}} + O(1)$ iterations.  A more careful argument using a Galton-Watson process gets the constant down to $\log_2 e$.
%FIXME: Explain this better.

\subsection{General LLL problems}
%FIXME

\subsection{The simple Moser-Tardos algorithm}
%FIXME

\subsection{A weaker condition}
Moser and Tardos actually prove that their algorithm terminates successfully in polynomial time under a weaker condition than $e p (d+1) < 1$.  Under some assignment of a weight to each flaw $x: \mathcal{A} \to (0,1)$, the following condition must hold for every flaw $A_i$:
\seqn
  P(A_i) \leq x(A_i) \prod_{j \in \Gamma(i)} (1-x(A_j)) (\forall A_i \in \mathcal{A}) \label{eqn:general-lll-req}
\eeqn

This condition is more complicated and unnecessary for exposition, so we have  focused mostly on the simpler one.  However, it is often the condition cited in the literature, so it is important to see the connection between the two conditions.  Let us briefly show that $e p (d+1) < 1$ implies the weaker condition with $x(A_i) = \frac{1}{d+1}$ for all $i$.  $P(A_i) \leq p$ for all $i$ by definition of $p$, so the (claimed) stronger condition gives us:
\seqn
  P(A_i) \leq p &< \frac{1}{d+1} \frac{1}{e}  \\
  &< \frac{1}{d+1} (1 - \frac{1}{d+1})^d \\
  &\leq \frac{1}{d+1} \prod_{j \in \Gamma(i)} (1-\frac{1}{d+1}) \\
  &= x(A_i) \prod_{j \in \Gamma(i)} (1 - x(A_j)) .
\eeqn
